---
title: "Introduction to `ggplot`"
author: "YOUR NAME HERE"
date: "Week 02"
output: html_notebook
---



# Week 02 - Flowers for Algernon

> Anyone who has common sense will remember that the bewilderments of the eye
 are of two kinds, and arise from two causes, either from coming out of the light
 or from going into the light, which is true of the mind's eye, quite as much as
 of the bodily eye.

Our data set this week is the `diamonds` data set. It contains the price and
attributes of nearly 54,000 round cut diamonds. Variables include:

- **price:** price in US dollars (\$326–\$18,823)
- **carat:** weight of the diamond (0.2–5.01)
- **cut:** quality of the cut (Fair, Good, Very Good, Premium, Ideal)
- **color:** diamond color, from D (best) to J (worst)
- **clarity:** a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2,
VS1, VVS2, VVS1, IF (best))
- **x:** length in mm (0–10.74)
- **y:** width in mm (0–58.9)
- **z:** depth in mm (0–31.8)
- **depth:** total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43–79)
- **table:** width of top of diamond relative to widest point (43–95)
 
You do not need to be knowledgeable about diamonds or their cut/value to succeed
on this lab. All you need is the data and a touch of common sense.

```{r echo=false, fig.align='center'}
url <- "https://imgs.xkcd.com/comics/scientific_paper_graph_quality.png"
knitr::include_graphics(url)
```



# Stories from Data

One of the most important skills you can learn is how to effectively understand
data visually. To do so, you must first understand when to use different kinds
of graphs/plots of data.

Below are some useful links which we discussed in class:
- [Data To Viz](https://www.data-to-viz.com/)
- [GGPlot Documentation](https://ggplot2.tidyverse.org/index.html)

To get help with ANY command in R, prepend the function name with a question
mark. For example, to get help with the `length` command, enter `?length` in the
R Console and RStudio will display the help documentation for that function. This
is a very useful feature.



# Lab

## Task 01

- Load the tidyverse and our data.
- Look at the output from this code chuck. How many rows of data are in the 
  diamonds data set?

```{r}
## ALWAYS load needed packages BEFORE trying to use them.
## The tidyverse gives us our graphing tools. 
## And readr gives us a good way to import CSV files.

library(tidyverse)
data(diamonds)
diamonds
``` 

Fun Fact: One year I asked students to complete this lab in Excel. It crashed
the VMs in the lab. Excel can do this, but it takes quite a bit of RAM.

## Task 02

 Create a scatter plot which shows the relationship between `carat` and `price`.

   - The first step is to use the `ggplot` function.
   - See `geom_point()`

The following code chunk won't work until you replace all the ?s. Note the red 
circles to the left of the line numbers. This is RStudio telling you the code 
has some problems. In this case, that's not an accident but it often is.

```{r}
ggplot(
  data = ?,
  aes(x = ?, y = ?)
) +
  geom_?()
```


## Task 03

Create a box plot which shows how `price` varies across diamonds of different
`clarity`.

```{r}
## YOUR CODE HERE

```


## Task 04

Create a bar plot to determine the most common diamond `cut` in this data set.

   - `geom_bar()` will be useful.
   
Question: What is the most common diamond `cut`? (You will need to know this for
Canvas.)

```{r}
## YOUR CODE HERE

```


## Task 05

Create a histogram of diamond price. Do you notice anything about the
distribution of diamond prices? How would you describe the distribution of
diamond prices? (The default number of bins, 30, is fine.)

   - See `geom_histogram`
   - [Normal Distribution - Wikipedia](https://en.wikipedia.org/wiki/Normal_distribution)
   - [Long Tail - Wikipedia](https://en.wikipedia.org/wiki/Long_tail)

```{r}
## YOUR CODE HERE

```


## Task 06

Create a second histogram of diamond price, and stratify this by diamond cut.
   - Use `facet_wrap()`. For help, `?facet_wrap` in the console.
   - Which cut has the least variance?
   - [Variance - Wikipedia](https://en.wikipedia.org/wiki/Variance)

```{r}
## YOUR CODE HERE

```


# Next Steps

Once done here, please complete the questions in Canvas, if you have not already
done so. Have a great week!
   
