---
title: "Introduction to `ggplot`"
author: "Andy Choens"
date: "Week 02"
output: 
  html_notebook:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: cerulean
---

# Tonight

- Learning Isn't Easy
- Data Life Cycle
- Last Week Tonight
    - A word on `library()`
    - Accessing a column
- Intro To Data Viz - GGPLOT

Read over this document. It will help you get oriented to the lab. Although you
don't need to complete the lab before class. But I do recommend looking at it.
If you come to class prepared, we can spend our class time doing fun stuff
like trying out your ideas, answering your questions, digging into data. Or, I
can lecture for an hour and a half and you can all take another nap.

You can read this documentation one of TWO ways:

1. As-is.
2. Hit the Preview button in the menu and render the discussion into a
   HTML document. (This is the pretty option.)



# Learning isn't easy

> The bad news is that when ever you learn a new skill your going to suck. It’s
going to be frustrating. The good news is that is typical and happens to
everyone and it is only temporary. You can’t go from knowing nothing to becoming
an expert without going through a period of great frustration and great
suckiness.
**--Hadley Wickham**

In addition to these lecture notes, labs, readings, videos, etc. there are other
ways to find help.

- [Stack Overflow](https://stackoverflow.com/questions/tagged/r)
- [Tidyverse Learn](https://www.tidyverse.org/learn/)
- `?mean` If you prepend a `?` to a command, R provide you with built-in 
  function documentation.
- [ggplot reference](https://ggplot2.tidyverse.org/reference/index.html#section-plot-basics)
- [Use Google Search](https://www.google.com/search?source=hp&ei=y-kNYLykM_-i5NoPo5m76AE&q=R+mean&oq=R+mean&gs_lcp=CgZwc3ktYWIQAzILCAAQsQMQgwEQyQMyAggAMgIIADICCAAyAggAMgIIADICCAAyAggAMgIIADICCAA6CAgAEOoCEI8BOggIABCxAxCDAToICC4QsQMQgwE6CAguEMcBEKMCOg4ILhCxAxCDARDHARCjAjoCCC46DgguELEDEIMBEMkDEJMCOggILhDHARCvAToFCAAQsQM6BQguELEDOgUIABDJA1DkFVifGmCmHGgBcAB4AIAB0gGIAegIkgEFMC40LjKYAQCgAQGqAQdnd3Mtd2l6sAEK&sclient=psy-ab&ved=0ahUKEwj8tcrlxLXuAhV_EVkFHaPMDh0Q4dUDCAg&uact=5)

Long story short, programming is not easy. And it can be frustrating when you 
are learning. I'd like to have a class forum, but we need to make this decision
together.

- Class Forum?
    - Yes/No
    - If Yes, How?

Let's discuss in class.

# Data Life Cycle

```{r echo=false, fig.align='center'}
url <- "https://d33wubrfki0l68.cloudfront.net/571b056757d68e6df81a3e3853f54d3c76ad6efc/32d37/diagrams/data-science.png"
knitr::include_graphics(url)
```

One of the high-level goals of this class is to help you become familiar with
this data life cycle.

- Import (CSV, Excel, web-scraping, API, databases)
- Munge (Tidy/Transform)
- Visualize/Model
- Communicate (Write a report, lab, etc.)

A traditional statistics course will do SOME data-viz and a bunch of modeling.

- **Good News:** Modeling is important!
- **Bad News:** Data is never clean or shaped the way you need it. And a good plot 
  is worth a thousand t-tests.
- Do you see how I'm using plain text to provide structure/emphasis?

**_Tonight, we will introduce data viz using ggplot._**



# Last Week Tonight

```{r echo=false, fig.align='center'}
url <- "https://imgs.xkcd.com/comics/drinking_fountains.png"
knitr::include_graphics(url)
```

## Library

- You will see `library(tidyverse)` at the top of nearly everything we do.
    - This is a package.
    - Similar to Chrome extensions.
- The tidyverse is actually a metapackage which contains a dozen or so packages.
    - These packages share a common style/work-flow and form the core of what
      we would consider "modern" R.
    - And it's easier than loading them all individually.
    - If you are following along at home, and none of my examples work, maybe
      you need to load the tidyverse.
    - This package includes `ggplot2` which is the most widely use data-viz
      package on R.

## Code Chunks

- Rmd files, also called RMarkdown files, contain plain text and computer code.
- The code usually lives in a thing called a code chunk. It is fenced like 
this:

```{r}
## THIS IS AN EMPTY CODE CHUNK!

```

## Some Data

- R comes with example data sets.
- Each column in a data frame is a vector.
- To learn more about this data set, `?chickwts`
- Let's take a look at what we have here.
- If the code below fails, make sure you have loaded the data by running this
  code chunk.

```{r eval=FALSE}
## The first command fails if you haven't loaded the tidyverse.

data("chickwts")
chickwts
?chickwts ## This opens the data documentation in the Help pane.
```

Now that's we've seen the data, let's plot it. Let's look at the distribution
of chick weight by feed type.

```{rm fig.align='center', eval=FALSE}
data("chickwts")

ggplot( 
  data = chickwts,
  aes(x = feed, y = weight)
  ) +
  geom_boxplot()
```

Oooh. I bet that returned an error.

- PLEASE REMEMBER WHAT THIS KIND OF AN ERROR LOOKS LIKE!!!!
- This error is telling you you haven't imported the tidyverse packages.
- Try running the NEXT code chunk. Can you spot the difference?

### Setup

This is the setup section for this lecture, even if it is waaay down here.

```{r fig.align='center', message=FALSE, warning=FALSE, paged.print=TRUE}
library(tidyverse)
data("chickwts")

ggplot( 
  data = chickwts,
  aes(x = feed, y = weight)
  ) +
  geom_boxplot()
```

- We will often need several packages.
- I load these at the top of the notebook in a code chunk called `setup`.
- You only need do this once per session.
- I will, however, often put them somewhere if I think it is possible you might
  have skipped the upper part of the lab/lecture.

## Data Frame Review
  
- You can use the `$` operator to access a column.

```{r}
chickwts$weight ## SO many different chicks
```

And you can apply a function to a column.

```{r}
mean(chickwts$weight)
sd(chickwts$weight)
```

You can even make  new data frame from this!

```{r}
d <- data.frame(
  avg = mean(chickwts$weight),
  sd = sd(chickwts$weight)
)
d
```

R is a functional programming language. This means we will apply functions 
(like `mean()`) to objects (like `chickwts`) to produce a result.

- Good function names are verbs. (Stats functions like `mean()` are exceptions.)
- Good object/data frame names are nouns.



# Intro to Data Viz

```{r echo=false, fig.align='center'}
url <- "https://imgs.xkcd.com/comics/data.png"
knitr::include_graphics(url)
```

## Data Visualization Skillz!

- Who can tell me about (why we us it, what kind of data to use it with):
    - histogram?
    - bar plot?
    - density plot?
    - scatter plot?
    - violin plot? (Yes, this is a thing.)
- Good tips: [https://www.data-to-viz.com/](https://www.data-to-viz.com/)
    
There are may more. But these are the basics. Please note that this list has 
more tools/options for continuous data than it does for categorical data.

## Class Survey!

```{r message=FALSE, warning=FALSE}
library(tidyverse)
ClassSurvey <- read_csv("includes/ClassSurvey.csv")
ClassSurvey
```
- Names have been abbreviated.
- I removed columns where you "wrote" something to me.
    - The goal here is not to embarrass anyone, but to have a simple/fun data
      set about all of you.
    - Is this scientific? Alas, no.
- We can/will use this to help learn about Data Viz using R.

## What year are you in at ACPHS?

- Select the option that best describes where you are in your educational
journey.
   - 1st Year (Freshman)
   - 2nd Year (Sophomore)
   - 3rd Year (Junior)
   - 4th Year (Senior)
   - 5+ (You like it here)
   - 999 (Grad School)

Before plotting:
- What kind of data are we plotting?
- What plot type is best for this data type?

```{r fig.align='center'}
## Two rows of code.
## And a plot.
ggplot(data = ClassSurvey, aes(x = WhatYear)) +
  geom_bar() ## geom_bar == bar plot!
```

The ggplot pattern:

- `ggplot(data = Foo, aes(x = x, y = y, color = bar))`: Creates the plot/space.
- `geom_something() +`: Defines what type of plot we want.
- `stuff()`: Helper definitions to define our plot.
- `facet_wrap(~c)`: Stratifies our results.

Example:

```{r fig.align='center'}
## Two rows of code.
## And a plot.
ggplot(data = ClassSurvey, aes(x = WhatYear)) +
  geom_bar() +
  facet_wrap(~DegreeProgram)  ## PLEASE NOTIVE THE USE OF THE TILDA!!!!
```

## Percent, not counts

- Sometimes, you don't want the different COUNTS of things, but the percent.
- The shape of the previous plot and the next plot are identical, but please
  see how the y-axis has a very different range.

```{r fig.align='center'}
## This is the pure ggplot way to do this.
## I rarely do it this way, but it does work.
ggplot(ClassSurvey, mapping = aes(x = WhatYear, y = 100*(..count..)/sum(..count..) )) +
  geom_bar(fill="darkred") +
  labs(y = "Percent", x="Year In School") +
  ylim(0,100)
```

## What is your current age?

- In reality, age is a continuous number.
- But we rarely capture it that way.
- We often discretize age, which is unwise, but that's another lesson.
- Age is a discrete numeric, so we should not use a box plot on it.
- And this looks a mess.

```{r fig.align='center'}
ggplot(ClassSurvey, aes(x = CurrentAge)) +
  geom_histogram(fill="darkred") + ## geom_histogram: HISTOGRAM!!!
  labs(title = "Student Age", x = "Current  Age", y = "N")
```

In our data, we didn't collect anyone's age as 19.2 or 19.8 although this is 
obviously true. So, although histogram is arguably the right type of plot, 
it doesn't actually work.

But, we can get a nice looking plot using the . . . bar plot!?!

```{r fig.align='center'}
ggplot(ClassSurvey, aes(x = as.factor(CurrentAge))) +
  geom_bar(fill="darkred") +
  labs(title = "Student Age", x = "Current  Age", y = "N")
```

## YOUR TURN

Do your classmates like math? Use the column `LikeMath` to answer this.

```{r fig.align='center'}
## YOUR CODE HERE!

```

- Now, let's compare your collective love of math across degree program.
- Stratify (`facet_wrap()`) by `DegreeProgram`.

```{r fig.align='center'}
## YOUR CODE HERE!

```

The truth is, there aren't very many questions I can easily ask on a "class 
survey" which will result in a truly continuous variable. I could ask you your
age in days or your height in inches (or centimeters). But most of our survey
data is categorical, which will be interesting on the lab.

## Plotting numerical chick data

We can use the `chickwts` data from before to better understand how a
histogram works.

```{r fig.align='center'}
ggplot(chickwts, aes(x = weight)) +
  geom_histogram(fill="darkblue") +
  labs(title = "Chick Weight Distribution", x = "Weight (gm)", y = "N")
```

- We can see that chick weight is _slightly_ bi-modal and long-tailed.
- Notice that R is asking us to manually set the number of bins in our plot.
- `geom_histogram` has a parameter called `bins`, which sets the number of bars
  in the histogram.
- This is different from a bar chart, where the number of bars is determined
  on the number of factors in the columns.
- How do we know what to set the bins to?

There are standards:

- [https://en.wikipedia.org/wiki/Histogram#Number_of_bins_and_width](https://en.wikipedia.org/wiki/Histogram#Number_of_bins_and_width)
- Square-root choice: `r sqrt(length(mpg$cty))`
this case.)
- [Sturges' formula](="https://wikimedia.org/api/rest_v1/media/math/render/svg/c020d823eb7a197d7db70bc66f8189c2ed7782b8"):
    - `r log2(length(mpg$cty))+1`
- With this many standards, you may be forgiven for thinking there is no
  standard.
- And you would be right. The question is really this - what are you trying to
  show about the distribution of the data?

But, there are other ways to show the distribution of a continuous variable
such as a density plot. (I like density plots.)
  
```{r fig.align='center'}
ggplot(chickwts, aes(x = weight)) +
  geom_density(color="darkblue") +
  labs(title = "Chick Weight Distribution", x = "Weight (gm)")
```

- This is the same basic graph, but it is shown using a density plot. 
- The area under the graph must be equal to 1.
- Notice that for a box plot, I used `fill` to change the color. For density 
  plots, use `color` instead.
- And if we want to go overboard, we can do both in one plot.

```{r fig.align='center'}
ggplot(chickwts, aes(x = weight)) +
  geom_histogram(aes(y = stat(density)), fill="darkblue", binwidth = 10) +
  geom_density(color = "black") +
  labs(title = "Chick Weight Distribution", x = "Weight (gm)", y = "N")
```

- [IMS: Histogram](https://openintro-ims.netlify.app/summarizing-visualizing-data.html?q=histo#histograms)

There are other ways we can try to understand the distribution of chick weight.
- box plots
- violin plots

```{r fig.align='center'}
ggplot(chickwts, aes(x = feed, y = weight)) +
  geom_boxplot(color="darkblue") +
  labs(title = "Chick Weight Distribution", x = "Feed Type", y = "Weight (gm)")
```

```{r fig.align='center'}
ggplot(chickwts, aes(x = feed, y = weight)) +
  geom_violin(color="darkblue") +
  labs(title = "Chick Weight Distribution", x = "Feed Type", y = "Weight (gm)")
```

If you aren't familiar with these:

- [IMS Box Plots](https://openintro-ims.netlify.app/summarizing-visualizing-data.html?q=box%20#box-plots-quartiles-and-the-median)
- Violin plots are not in IMS, and we can talk about them in class.

## Graphing continuous variables

We will switch to another data set now, the mpg data.

```{r}
data(mpg)
mpg
?mpg
```

A scatter plot let's us show the RELATIONSHIP between two continuous variable.

```{r fig.align='center'}
## This is slightly different from what is shown in the book.
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = manufacturer)) + 
  geom_point()
```

- What do we call the variable on the X Axis?
- What do we call the variable on the Y Axis?
- What are we implying about these two numbers?
- [IMS Scatter Plots](https://openintro-ims.netlify.app/summarizing-visualizing-data.html?q=scatt#scatterplots)

And we can even add a regression line.

```{r fig.align='center'}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
   geom_point(color = "steelblue", size = 2) +
   geom_smooth(method = "lm", color="maroon", size = 1.5)
```

Ooooh. Pretty colors.

## Facets

When we want to stratify our results, we can use facet_wrap, again.

```{r fig.align='center'}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
   geom_point(color = "steelblue", size = 2) +
   geom_smooth(method = "lm", color="maroon", size = 1.5) +
   facet_wrap(~class, nrow = 2)
```

## Correlation Plot

```{r fig.align='center'}
## Slightly tougher
library(GGally) ## Yet Another Package
ggcorr(mpg, label = TRUE)
```

- This is not in the book.
- But it is useful.
- We need to unpack this a little. First, what is this telling us and how might
  it be useful?

```{r}
cor(mpg$cty, mpg$hwy)
```



# Questions?

```{r echo=false, fig.align='center'}
url <- "./includes/ggplot-logo.png"
knitr::include_graphics(url)
```

